ExpressionView - an interactive viewer for biclusters identified in gene expression data

### 2010-01-11 expressionviewnote/expressionview.tex
More expressionview paper updates, sent to co-authors.

\usepackage[dvipdfm,colorlinks,bookmarks=false,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
\newcommand{\thefirstpage}{1} \newcommand{\thelastpage}{1}

\copyrightyear{2009}
\pubyear{2009}

@@

\section{Availability:} 
\href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{R package with open source license}

\section{Contact:} 
\href{andreas.luescher@a3.epfl.ch}{Andreas L\"uscher}

\section{Supplementary information:}
Screenshots, tutorials and sample data sets can be found on the ExpressionView \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.

@@

\section{Package design and workflow}
The ExpressionView package contains two independent parts: An ordering algorithm implemented in C++ and a Adobe Flash applet written in ActionScript and Adobe Flex. Fig.~\ref{fig:workflow} schematically summarises the typical workflow that uses the genomic data available as a Bioconductor ExpressionSet and combines it with the biclustering results to produce an ExpressionView data file that can be explored with the Flash applet. In the following, we describe the different steps in more detail.
\begin{figure*}[!tpb]
\centerline{\includegraphics[width=0.9\linewidth]{fig1-crop}}
\caption{How to use ExpressionView: Starting from gene expression data in the form of a Bioconductor ExpressionSet, the user first runs a biclustering algorithm to determine co-regulated groups of genes and samples. In a second step, the rows and columns of the gene expression matrix are rearranged to produce an easily readable overview of the results. The last step consist in combining the gene expression data and its associated metadata with the results from the biclustering and produce an ExpressionView data file that can be explored with the interactive Flash applet.}\label{fig:workflow}
\end{figure*}
For users who prefer to install the viewer as a standalone program, we also provide a platform independent Adobe AIR version that can be downloaded from the \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}.

\subsection{Gene expression data and biclusters}
ExpressionView is designed to work with gene expression data in the form of a Bioconductor ExpressionSet. This class provides a user-friendly way to access the actual gene expression matrix and its associated metadata. There is a variety of biclustering algorithms described in the literature~\citep{madeira04,prelic06}, several of which have been implemented as R packages. ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03}%,csardi09
and the methods available in the Biclust package~\citep{kaiser08}. Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.

@@

\subsection{Export}
Once the optimal order is determined, the program rearranges the gene expression matrix accordingly and exports all the relevant information into an XML file. This format has the advantage of being self-explanatory and extensible. In addition, the files can be read and edited with any text editor. The actual gene expression data is rounded to two digit precision and encoded in Base64. The structure of the XML file is described in more detail on the ExpressionView \href{http://www2.unil.ch/cbg/index.php?title=ExpressionView}{website}. In most cases, these files are of the order of a few Megabytes and can thus easily be sent to co-workers.

>>>>

\usepackage[colorlinks,citecolor=blue,linkcolor=red,urlcolor=blue]{hyperref}
\newcommand{\thefirstpage}{1} \newcommand{\thelastpage}{1}

\copyrightyear{2010}
\pubyear{2010}

@@

\section{Availability:} 
\href{http://www.unil.ch/cbg/ExpressionView}{R
  package with open source license, \url{http://www.unil.ch/cbg/ExpressionView}}

\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}

\section{Supplementary information:}
Screenshots, tutorials and sample data sets can be found on the ExpressionView \href{http://www.unil.ch/cbg/ExpressionView}{website}.

@@

\section{Package design and workflow}
The ExpressionView package contains two independent parts: An ordering algorithm implemented in C++ and a Adobe Flash applet written in ActionScript and Adobe Flex. Fig.~\ref{fig:workflow} schematically summarises the typical workflow that uses the genomic data available as a Bioconductor ExpressionSet and combines it with the biclustering results to produce an ExpressionView data file that can be explored with the Flash applet. In the following, we describe the different steps in more detail.
\begin{figure*}[!tpb]
\centerline{\includegraphics[width=0.9\linewidth]{fig1-crop}}
\caption{How to use ExpressionView: Starting from gene expression data in the form of a Bioconductor ExpressionSet, the user first runs a biclustering algorithm to determine co-regulated groups of genes and samples. In a second step, the rows and columns of the gene expression matrix are rearranged to produce an easily readable overview of the results. The last step consist in combining the gene expression data and its associated metadata with the results from the biclustering and produce an ExpressionView data file that can be explored with the interactive Flash applet.}\label{fig:workflow}
\end{figure*}
For users who prefer to install the viewer as a standalone program, we also provide a platform independent Adobe AIR version that can be downloaded from the \href{http://www.unil.ch/cbg/ExpressionView}{website}.

\subsection{Gene expression data and biclusters}
ExpressionView is designed to work with gene expression data in the form of a Bioconductor ExpressionSet. This class provides a user-friendly way to access the actual gene expression matrix and its associated metadata. There is a variety of biclustering algorithms described in the literature~\citep{madeira04,prelic06}, several of which have been implemented as R packages. ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03} %,csardi09
and the methods available in the Biclust package~\citep{kaiser08}. Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.

@@

\subsection{Export}
Once the optimal order is determined, the program rearranges the gene expression matrix accordingly and exports all the relevant information into an XML file. This format has the advantage of being self-explanatory and extensible. In addition, the files can be read and edited with any text editor. The actual gene expression data is rounded to two digit precision and encoded in Base64. The structure of the XML file is described in more detail on the ExpressionView \href{http://www.unil.ch/cbg/ExpressionView}{website}. In most cases, these files are of the order of a few Megabytes and can thus easily be sent to co-workers.
